---
source: Rmd
title: "Common Analyses"
teaching: 10
exercises: 2
questions:
- "What are the most common single cell RNA-Seq analyses?"
objectives:
- "Explain how to use RMarkdown with the new lesson template."
- "Demonstrate how to include pieces of code, figures, and challenges."
keypoints:
- "Edit the .Rmd files not the .md files"
- "Run `make serve` to knit documents and preview lesson website locally"
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")

library(tidyverse)
library(Seurat)

data_dir <- '../data'
```

## Read Data from Previous Lesson

```{r load_data}
liver <- readRDS(file.path(data_dir, 'lesson04.rds'))
```

## Normalization (log and more specialized) 

Instead of working with raw count data measured across cells
that were sequenced to highly variable depths, we conduct
normalization to try to make gene expression values follow
a more stable distribution as well as being more comparable
between cells.

### Log Normalization

The count data is usually log-normally distributed. 
Many statistical methods work best when the data is normally distributed. 
We also would like to correct for variability in sequencing depth 
between cells, the nature of which is purely technical.
Log normalization will give us normalized gene expression which represents
the log of the number of counts per 10,000 reads.

```{r normalization, message=FALSE}
liver <- liver %>%
              NormalizeData(normalization.method = "LogNormalize")
```

### Finding Variable Features

Next we will find a subset of features showing high cell-to-cell variation 
in the dataset (that is, they are highly expressed in some cells and lowly
expressed in others). 

```{r var_features, message=FALSE}
liver <- liver %>% 
              FindVariableFeatures(nfeatures = 2000)

# Identify the 25 most highly variable genes
top25 <- head(VariableFeatures(liver), 25)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(liver)
plot2 <- LabelPoints(plot = plot1, points = top25, xnudge = 0, 
                     ynudge = 0,repel = TRUE)
plot1 + plot2
```

## Cell cycle assignment 

We will also show how to predict cell cycle state.
This approach is outlined in the Seurat vignette at
[this link](https://satijalab.org/seurat/articles/cell_cycle_vignette.html).


```{r cellcycle}
cc.genes <- readLines(file.path(data_dir,
  'regev_lab_cell_cycle_genes_mm.fixed.txt'))
s.genes <- cc.genes[1:43]
g2m.genes <- cc.genes[44:98]

liver <- CellCycleScoring(liver, s.features=s.genes, 
  g2m.features=g2m.genes, set.ident=FALSE)
```

Seurat will provide a quantitative estimate of the cell's chance of being
in different phases of the cell cycle `S.Score` and `G2M.Score`, as well as
a categorical prediction of which phase the cell is in 
(`Phase` -- G1, G2M, S).

### Scale Data

Now we apply a linear transformation that is often used in initial 
scRNA-Seq processing. This transformation standardizes the expression of
each gene, setting the mean across cells to 0 and variance to 1. 
This helps all genes to contribute to the inferred variability rather
than just the highly-expressed genes.

We will "regress out" the signals of technical confounders including
%MT and the number of UMIs.

```{r scaling, message=FALSE}
liver <- liver %>%
              ScaleData(vars.to.regress = c("percent.mt", "nCount_RNA"))
```

### Principal Component Analysis

Next we reduce the dimensionality of the data. You have probably heard
of PCA as a technique for summarizing major axes of variation in a dataset.
Here, we perform PCA on the single cell gene expression data in order to
place each cell in a multidimensional space with lower dimension (say 20-40)
than the complete expression matrix (~20,000 genes).

```{r PCA, message=FALSE}
liver <- liver %>%
              RunPCA(verbose = FALSE, npcs = 100)
```

It is usually not very useful to view the raw PCs themselves:
```{r pcplot}
DimPlot(liver, reduction = "pca")
```

Instead we will take some of the PCs and use them for a further
summarization of the data. Namely, we will use the PCs as input to the
UMAP (or t-SNE) algorithm which projects our cells onto a 2D space
(the computer screen). 

A significant challenge in scRNA-Seq is deciding how many PCs to use.
You can think of each PC as capturing pathway-level transcriptional variation
across the cells in your dataset. Thus even if the transcriptome is sparse
within each cell (it is), you can still compare different cells across major
axes of variation. We don't want to use too few PCs, otherwise we 
might miss significant axes of variation (e.g. a cell subtype or minor cell 
type). We also don't want to use too many PCs since, as you go out in PC-space,
the PCs increasingly represent more noise and less biological reality.

We will use a very simple method to choose the number of PCs: the elbow
method. Using this method we look for where the elbow plot stops dropping
precipitously.
```{r elbow}
ElbowPlot(liver, ndims = 100)
```

Let's zoom in more and see what things like under 50 PCs.
```{r elbow2}
ElbowPlot(liver, ndims = 50)
```

We would say that the standard deviation in PCs really starts to stablize
around N = 27 PCs. Let's use this value moving forward.

```{r elbow2}
num_pc <- 27
ElbowPlot(liver, ndims = 40) + geom_vline(xintercept = num_pc)
```

## Dimensionality reduction (UMAP, tSNE, etc) 

As mentioned above, dimensionality reduction allows you to actually 
visualize your data! The two methods below are widely used in the
single cell community.

> Uniform Manifold Approximation and Projection (UMAP) [van der Maaten & Hinton, 2008](https://www.jmlr.org/papers/volume9/vandermaaten08a/vandermaaten08a.pdf).

> t-Distributed Stochastic Neighbor Embedding (t-SNE) [McUnnes et al](https://arxiv.org/abs/1802.03426) 

These methods generally do a very effective job of putting similar points near each other in the reduced-dimensionality space. Thus cells from 
the same clusters are likely to be placed in the same region of the UMAP/t-SNE.

UMAP is more widely used the t-SNE at the current time.
Note that you should caution yourself not to overinterpret UMAP plots.
Although UMAP does optimize both local and global similarity for points
being projected onto a 2D space, UMAP contains no guarantee that similar points
must be near each other.

```{r umap}
liver <- RunUMAP(liver, reduction = 'pca', dims = 1:num_pc, 
    verbose = FALSE)
```

## Clustering 

```{r seurat3, message=FALSE}

liver <- FindNeighbors(liver, reduction = 'pca', 
                       dims = 1:num_pc, verbose = FALSE) %>%
           FindClusters(verbose = FALSE, resolution = 0.3)
UMAPPlot(liver, label = TRUE, label.size = 6)
```

## Finding marker genes 

Now we will find marker genes for our clusters. Finding marker genes takes a
while so we will downsample our data to speed up the process.
Even still this may take a few minutes.

```{r markers}
liver_mini <- subset(liver, downsample = 300)
markers <- FindAllMarkers(liver_mini, only.pos = TRUE, 
    logfc.threshold	= log2(1.25), min.pct = 0.2) 
```


<!-- ## Annotating cell types (+ automated options e.g. SingleR)  -->
<!-- DAS - I don't think these methods are generally all that useful. The -->
<!-- cell types they get right are usually easy to identify, while the -->
<!-- harder types are not identified correctly by SingleR. -->
<!-- Should we include this as opinion/narrator's comment? -->

We will again use the Seurat object in the next lesson. Save it now and we will 
load it at the beginning of the next lesson. 

```{r save_seurat}
saveRDS(liver, markers, file = file.path(data_dir, 'lesson05.RData'))
```


## Session Info

```{r session_info}
sessionInfo()
```
